\texttt{Davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
fix version strings in metadata table
implement @jeremymanning's suggestions -- thanks!

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata value} \\
\hline
C1 & Current code version &  v0.2.4 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.2.4} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, ipykernel, PyZMQ.\newline Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments, and
     dependencies & Dependencies:~Python $\geq 3.6$, packaging, setuptools.\newline Supported OSes: MacOS, Linux, Unix-like.\newline Supported IPython environments: Jupyter Notebooks, JupyterLab, Google Colaboratory, Binder, Kaggle, IDE-based notebook editors, IPython shell. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & \href{mailto:contextualdynamics@gmail.com}{contextualdynamics@gmail.com} \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

We designed \texttt{Davos} for use in research settings, where code for numerous different tasks---from processing data, to running statistical analyses, to generating figures and tables for publication---is frequently shared among collaborators while working on a project, and eventually with the broader scientific community and general public upon a its completion.
In these contexts, ensuring that shared code yields consistent, reproducible outputs across users and over time is crucially important, yet the tools available to researchers for doing so can be complex to set up and challenging to properly use.
This has the dual effect of discouraging scientists from sharing their code in a reproducible way (or at all), and making it significantly harder (or impossible) for others to successfully reproduce their results, adopt or extend their methods, and contribute to or build upon their work.
Ultimately, the need to install and master additional tools in order to share and run reproducible code can impede progress both on the individual level and on the broader scientific level.

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as virtual environments, containers, and virtual machines (see Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment, \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to virtual environments as it similarly provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

@@

Third, running a shared notebook that uses \texttt{Davos} to manage its dependencies often requires less (but never more) space on the user's system than running an identical notebook inside a virtual environment.
While typical virtual environment directories will contain \textit{all} requirements listed in their configuration files, \texttt{Davos}'s isolated project directories (by default; see Sec.~\ref{subsec:projects}) contain only those not already available in the user's existing Python environment.
This is another feature made possible by \texttt{Davos}'s runtime-based dependency management scheme.
\texttt{Davos} can safely draw from packages in from the user's main environment to satisfy a notebook's dependencies, when possible, because if those packages were to be updated or removed at any point such that they no longer met the notebook's requirements, appropriate versions could (and would) be installed into the project directory the next time the notebook was executed.

Finally, neither virtual environments nor \texttt{Davos}-enhanced notebooks can bundle specific versions of Python with which to run shared code.
However, \texttt{Davos}'s \texttt{require\_\-python()} function provides a simple mechanism for indicating a required or constrained Python version and alerting the user at runtime (by raising an exception) if their Python version is incompatible.
In terms of \texttt{Davos}'s ability to ensure that shared code can be executed reproducibly by other users, this falls short of the capabilities of more complex tools that can provide complete copies of Python (see Fig.~\ref{fig:code-sharing}) and should be preferred over \texttt{Davos} in cases where this is necessary.
However, the functionality it provides over what is possible with a standard virtual environment is removing the \textit{expectation} that running a particular notebook will reproduce an expected outcome in situations where this is either impossible or not guaranteed.
With this understanding, a user may choose to install a compatible Python version through some other means or elect to still run the code, but will not be surprised by a potential failure to execute successfully or output an expected result.

@@

A second more subtle impact of our work is in providing a
proof-of-concept of how the ability to add new ``keyword-like''
operators to the Python language could be specifically useful to
researchers. With \texttt{Davos}, we accomplish this by leveraging
IPython notebooks' internal code parsing and execution machinery. We
note that, while other popular packages similarly use these mechanisms
to providing notebook-specific functionality (e.g.,
\cite{Hunt07,HeusEtal18}), this approach also has the potential to be
exploited for more nefarious purposes. For example, a malicious user
could design a Python package that, when imported, substantially
changes the notebook's functionality by adding new \textit{unexpected}
keyword-like objects (e.g., based around common typos). We also note
that this implementation approach means \texttt{Davos}'s functionality
is currently restricted to IPython notebook environments. However,
there have been early-stage discussions of providing this sort of
syntactic customizability as a core feature of the Python language,
including a draft proposal~\cite{Shan20}. In addition to enabling
\texttt{Davos} to be extended for use outside of notebooks, this could
lead to exciting new tools that, like \texttt{Davos}, extend the
Python language in useful and more secure ways.

@@

The \texttt{Davos} package supports reproducible research by providing
a novel, lightweight system for sharing notebook-based code. It
expands on Python's ``batteries included'' philosophy to enable running shared
notebooks with no setup required, and defines a simple, self-documenting format for
specifying dependencies in the spirit of ``literate programming'' paradigm Jupyter notebooks promote.
But perhaps the most exciting uses of the \texttt{Davos} package are those
that we have \textit{not} yet considered or imagined. We hope that the
research and scientific Python communities will find \texttt{Davos} to provide a convenient
means of managing project dependencies to facilitate code sharing and collaboration. We
also hope that some of the more advanced applications of our package
might lead to new insights or discoveries.

>>>>

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata value} \\
\hline
C1 & Current code version &  v0.2.3 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.2.3} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, ipykernel, PyZMQ.\newline Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments, and
     dependencies & Dependencies:~Python $\geq 3.6$, packaging, setuptools.\newline Supported OSes: MacOS, Linux, Unix-like.\newline Supported IPython environments: Jupyter Notebooks, JupyterLab, Google Colaboratory, Binder, Kaggle, IDE-based notebook editors, IPython shell. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & \href{mailto:contextualdynamics@gmail.com}{contextualdynamics@gmail.com} \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

We designed \texttt{Davos} for use in research settings, where code for numerous different tasks---from processing data, to running statistical analyses, to generating figures and tables for publication---is frequently shared between collaborators while working on a project, and eventually with the broader scientific community and general public upon its completion.
In these contexts, ensuring that shared code yields consistent, reproducible outputs across users and over time is critical, yet the tools available to researchers for doing so can be complex to set up and challenging to properly use.
This has the dual effect of discouraging scientists from sharing their code in a reproducible way (or at all), and making it significantly harder (or impossible) for others to successfully reproduce their results, adopt or extend their methods, and contribute to or build upon their work.
Ultimately, the need to install and master additional tools in order to share and run reproducible code can impede progress both on the individual level and on the broader scientific level.

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as virtual environments, containers, and virtual machines (Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment(s), \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to virtual environments in that it provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

@@

Third, running a shared notebook that uses \texttt{Davos} to manage its dependencies often requires less (but never more) space on the user's system than running an identical notebook inside a virtual environment.
While typical virtual environment directories will contain \textit{all} requirements listed in their configuration files, \texttt{Davos}'s isolated project directories (by default; see Sec.~\ref{subsec:projects}) contain only those not already available in the user's existing Python environment.
This is another feature made possible by \texttt{Davos}'s runtime-based dependency management system.
\texttt{Davos} can safely draw from packages in the user's main environment to satisfy a notebook's dependencies, when possible, because if those packages were to be updated or removed at any point such that they no longer met the notebook's requirements, appropriate versions would be installed into the project directory the next time the notebook was executed.

Finally, neither virtual environments nor \texttt{Davos}-enhanced notebooks can bundle specific versions of Python with which to run shared code.
However, \texttt{Davos}'s \texttt{require\_\-python()} function provides a simple mechanism for indicating a required or constrained Python version and alerting the user at runtime (by raising an exception) if their Python version is incompatible.
In terms of \texttt{Davos}'s ability to ensure that shared code can be executed reproducibly by other users, this falls short of the capabilities of more complex tools that can provide complete copies of Python (Fig.~\ref{fig:code-sharing}).
However, the functionality \texttt{Davos} provides over what is possible with a standard virtual environment is to remove the \textit{expectation} that running a particular notebook will reproduce an expected outcome in situations where this is either impossible or not guaranteed.
With this understanding, a user may choose to install a compatible Python version through some other means or elect to still run the code, but will not be surprised by a potential failure to execute successfully or output an expected result.

@@

A second more subtle impact of our work is in providing a
proof-of-concept of how the ability to add new ``keyword-like''
operators to the Python language could be specifically useful to
researchers. With \texttt{Davos}, we accomplish this by leveraging
IPython notebooks' internal code parsing and execution machinery. We
note that, while other popular packages similarly use these mechanisms
to providing notebook-specific functionality (e.g.,
\cite{Hunt07,HeusEtal18}), this approach also has the potential to be
exploited for more nefarious purposes. For example, a malicious user
could design a Python package that, when imported, substantially
changes the notebook's functionality by adding new \textit{unexpected}
keyword-like objects (e.g., based around common typos). We also note
that this implementation approach means \texttt{Davos}'s functionality
is currently restricted to IPython notebook environments. However,
there have been early-stage discussions of providing this sort of
syntactic customizability as a core feature of the Python language itself,
including a draft proposal~\cite{Shan20}. In addition to enabling
\texttt{Davos} to be extended for use outside of notebooks, this could
lead to exciting new tools that, like \texttt{Davos}, extend the
Python language in useful and more secure ways.

@@

The \texttt{Davos} package supports reproducible research by providing
a novel, lightweight system for sharing notebook-based code. It
expands on Python's ``batteries included'' philosophy to enable running shared
notebooks with no setup required, and defines a simple, self-documenting format for
specifying dependencies in the spirit of the ``literate programming'' paradigm Jupyter notebooks promote.
But perhaps the most exciting uses of the \texttt{Davos} package are those
that we have \textit{not} yet considered or imagined. We hope that the
research and scientific Python communities will find \texttt{Davos} to provide a convenient
means of managing project dependencies to facilitate code sharing and collaboration. We
also hope that some of the more advanced applications of our package
might lead to new insights or discoveries.
